---
title: "Tidying and Transforming Data"
author: "Alain T Kuiete"
date: "9/26/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tyding and Transformation

```{r}
library(readr)
library(stringr)
library(tidyr)
library(dplyr)
```

#### Importing csv file from my github
```{r}
airline <- read.csv("https://raw.githubusercontent.com/AlainKuiete/DATA607/master/arrival_delays.csv")
```

#### removing the blank row
```{r}
arr.delay <- airline[-3,]
row.names(arr.delay)<-c(1:4)
airline
```

#### Suppressing comma in numeric data
```{r}
arr.delay$Seattle <- gsub(",", "",arr.delay$Seattle)
arr.delay$Phoenix <- gsub(",", "",arr.delay$Phoenix)
```

#### Filling the blank cells
```{r}
arr.delay$X[2] <- arr.delay$X[1]
arr.delay$X[4] <- arr.delay$X[3]
arr.delay
```



#### Putting the appropriate variables. Los.Angeles, Phoenix,... at the header of columns are values of the variable airline.
#### First transformatiom  switch column to row and add the intermediate count column
```{r}
arr.delay <- gather(arr.delay,"Destination", "Count", 3:7)
```

#### Second Transformation split the (on time - delayed)column in two column.
```{r}
arr.delay <- spread(arr.delay, X.1, Count)
colnames(arr.delay) <- c("Airline", "Destination", "Delayed", "OnTime")

```
 

#### Converting columns Delayed and OnTime to numeric
```{r}
arr.delay$Delayed <- as.numeric(arr.delay$Delayed) 
arr.delay$OnTime <- as.numeric(arr.delay$OnTime)
arr.delay
```


#### Rate of Arrival delays per flight for Alaska
```{r}
AD <- sum(arr.delay$Delayed[which(arr.delay$Airline == "ALASKA")])
AO <- sum(arr.delay$OnTime[which(arr.delay$Airline == "ALASKA")])
RA <- AD/(AO + AD)
RA
```

### Rate of Arrival delays per flight for AM WEST
```{r}
WD <- sum(arr.delay$Delayed[which(arr.delay$Airline == "AM WEST")])
WO <- sum(arr.delay$OnTime[which(arr.delay$Airline == "AM WEST")])
RW <- WD/(WO + WD)
RW
```
#### Alaska rate of arriving delayed (13.3%) is higher than AM West rate of arriving delayed (10.9%)










