---
title: "DATA 607 ASSIGNMENT 1 -  LOADING INTO A DATA FRAME"
author: "Alain T Kuiete"
date: "8/31/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Mushrooms Dataset
## Exploring and inspecting the mushrooms dataset
```{r}
mushroom <- read.table('https://raw.githubusercontent.com/AlainKuiete/DATA607/master/agaricus-lepiota.data', sep = ",")
```

Looking at the head of mushrooms dataset
```{r}
head(mushroom)
```

The tail
```{r}
tail(mushroom)
```

The summary of the dataset
```{r}
summary(mushroom)
```

The structure of the mushrooms dataset
```{r}
str(mushroom)
```

## Subsetting the mushroom dataset
```{r}
md <- data.frame(is.consm = mushroom$V1, 
                 cap.chape = mushroom$V2, 
                 cap.color = mushroom$V4, 
                 odor = mushroom$V6, 
                 ring.type = mushroom$V20,
                 population = mushroom$V22)
```

## Structure of the new dataset
```{r}
str(md)
```
## Summary of the subset
```{r}
summary(md)
```


## Changing the abreviation inside the columns

```{r}
levels(md$is.consm) <- list(edible = "e", poisonous = "p")

levels(md$cap.chape) <- list(bell = "b", conical = "c", convex = "x", flat ="f", knobbed = "k", sunken="s" )

levels(md$cap.color) <- list(brown ="n", buff = "b", cinnamon = "c", gray = "g", green = "r", pink = "p",  purple = "u", red = "e", white = "w", yellow = "y" )

levels(md$odor) <-list(almond = "a", anise = "l", creosote = "c", fishy = "y", foul = "f", musty = "m", none = "n", pungent = "p", spicy = "s" )

levels(md$ring.type) <- list(cobwebby = "c", evanescent = "e", flaring = "f", large = "l", none = "n", pendant = "p", sheathing = "s", zone = "z")

levels(md$population) <- list(abundant = "a", clustered = "c", numerous = "n", scattered = "s", several = "v", solitary = "y" )

```
```{r}
str(md)
```

## Summary of the new dataset

```{r}
summary(md)
```

## Create a bar graph


```{r}
library(ggplot2)
qplot(md$cap.color, main = "Mushroom Cap color Attribute",
      xlab = "Cap color", ylab = "Count",
      color = I("red"))
```

## Mushrooms are most likely to be brown, gray, red, white or yellow.
